---
title: "Workshop Information"
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
```

## Prerequisites

### Bring a laptop that is correctly set up for R package development

Welcome to the R package development workshop. This will be a hands on workshop so participants are expected to **bring a laptop**. 

Before the course starts, please make sure that you have the most up to date version of R and Rstudio installed on your computer. Then, follow the instructions on the link below for your operating system to make sure you have the correct software needed to build R packages: 

https://support.rstudio.com/hc/en-us/articles/200486498-Package-Development-Prerequisites

## Time

#### Wednesday 27th November to Friday 29th November 2019 

### Daily Schedule (Tentative)

```{r}
Time <- seq(9, 16, by = .5)
Y <- -Time + 16
names(Y) <- Time
Day <- c("Wednesday", "Thursday", "Friday")
sched <- expand.grid(Time, Day)
colnames(sched) <- c("Time", "Day")
sched$What <- "Hands on"
sched$What[sched$Time %in% c(11.5, 14)] <- "Lecture"
sched$What[sched$Time %in% c(9, 14.5)] <- "Fika"
sched$What[sched$Time %in% c(12, 12.5)] <- "Lunch (provided)"

sched$Y <- Y[as.character(sched$Time)]
ggplot(sched, aes(x = Day, y = Y, fill = What)) + geom_tile(color = "grey80") + theme_minimal() + 
    scale_y_continuous("Time", breaks = Y + .25, labels = Time) + scale_fill_brewer(type = "qual") + 
    theme(legend.position = "bottom")
```

## Location

#### Nobels väg 12A, Campus Solna

```{r}
library(leaflet)
leaflet() %>%
  addTiles %>% # Add default OpenStreetMap map tiles
  setView(lng = 18.02504, lat = 59.34832, zoom = 16) %>%
    addPopups(lng = 18.02504, lat = 59.34832, "Workshop location (MEB)")
```


## Contact Information: 

- Michael Sachs (e-mail: michael.sachs@ki.se), Department of Medicine, Solna, Clinical Epidemiology Unit (KEP), Karolinska Institutet
- Alexander Ploner (e-mail: alexander.ploner@ki.se), Department of Medical Epidemology and Biostatistics (MEB), Karolinska Institutet

