---
title: "Publishing and Promoting a Package"
format: 
    revealjs:
      scrollable: true
execute:
  echo: true
  eval: false
---


## Getting Academic Credit for Your Work

Unfortunately you need to publish a paper not just a package to get credit, although times are changing! 

One option is software journals:
- The R Journal 
- Journal of Statistical Software 
- Journal of Computational and Graphical Statitics
- Journal of Open Source Software

These are all options, and they have different formats and requirements. 

## The R Journal 

The papers are basically a well-polished, extended vingette

- worked through examples, data should be be included in the package 
- give examples of all functions in the package and discuss options 
- discuss details of implementation

[pseval paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5774631/)

## Journal of Statistical Software

Actually has a pretty high impact factor! 

- More academic than The R J
- Interesting or well-written vignettes are perfect
- More focus on the methodology

[plotROC paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6347406/)

## Journal of Computational and Graphical Statistics

Old school stats journal

- Part of the AMSTAT family
- Focus is on the statistical/computational method
- Having a R package is just a bonus

## Journal of open source software

New kid on the block

- Focus is on the software, not the paper
- Short vignettes, extended descriptions
- Peer-review main focus is on the package itself, quality of documentation and code structure

## The other option is a case study 

- Write an application, practice or case study paper 
- Using your package for a interesting dataset or scientific problem
- Submit to a regular journal in your field

Many of the epidemiology journals will gladly accept descriptions of statistical software

## If all else fails

Self-publish!

```{r}
usethis::use_pkgdown()
```

If you have a nice README.md, and vignettes as html files, it is trivial to turn it into a nice-looking webpage. 

## CITATION file

If you do publish, make sure users know how to cite you: 

```{r}
usethis::use_citation()
```

Then people can find the citation using

```{r}
citation("packagename")
```


## Software review

<https://smartbiomed.github.io/software-review>


# Thank you!

