---
title: "Documentation"
author: "[Alexander Ploner](https://staff.ki.se/people/aleplo) - [Medical Epidemiology & Biostatistics](https://ki.se/en/meb), [Karolinska Institutet](https://ki.se/en)"
date: "2019-11-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, prompt = TRUE, comment = NA, collapse = TRUE)
```

# The R help system

## First line support: `?`

One of the central differences between scripts and packages is the availability of structured documentation integrated into the R commandline. Let's look at the help for a small function via `?abs` or `help(abs)`.

This gives you a detailed description of two straightforward arithmetic functions, with a reference, links to related subjects, and some examples that can be run at the command line via `example(abs)`.

Writing this documentation is (mostly) your responsibility. The standard format for writing help is the _R documentation_ format. These `.Rd`-files live in the `man`-folder of the source package directory, and R builds boths its interactive HTML help and the package reference files in `.pdf` format as seen on CRAN.

This is what the `.Rd` file for `abs` looks like: `MathFun.Rd`

The `.Rd`-format uses markup similar to LaTeX to define fields (`name`, `title`) and sections (`description`), and allows for lists, links and some formatting. 

## Building help from `.Rd` 

Let's start with a stupid function:

```{r}
say_hello = function(name = "my little friend", punct = ".")
{
  cat( paste0("Hello, ", name, punct), "\n", sep="")
}
say_hello()
```

Let's build a simple package in the working directory:
```{r, eval = FALSE}
library(devtools)
create("say")
wd("say")
dump("say_hello", file = "R/say.R")
rm(say_hello)  ## remove local copy
load_all()     ## load the package
say_hello()    ## this is the package-version
```

We can use a standard R function to generate a template for a `.Rd` file:

```{r, eval = FALSE}
dir.create("man")
prompt(say_hello, filename = "man/say_hello.Rd")
```

This is what the template thus generated looks like: `say_hello_template.Rd`

We can now edit the template to fit our needs. This is what the edited template could look like: `say.Rd`.

Note: while editing a `.Rd`file in RStudio, the usual _Compile report_ shortcut (`Ctrl-Shift-K`) will show a preview of the help page.

Once we are happy with the documentation (for now, at least), we can build the package again and check what it looks like:
```{r, eval = FALSE}
load_all()
?say_hello
```

Note: `%` is a comment character and can break the build. Here in the URL, the `%` is escaped by a backslash `\`

Note: links to other functions from the _See Also_ section do not work when the package is only loaded for development - the links will work when the package is installed:
```{r, eval = FALSE}
install()
?say_hello   ## link to paste should be active
```


## Easier: building help with roxygen2

Problems with writing `.Rd` files:

1. The markup is heavy, ugly and not very intuitive 
2. Documentation separate from code, hard to keep synced

Solution: package `roxygen2`

* use special comments + tags directly in source `.R` file
* write documentation right on top of function
* use `document()` to translate to `.Rd` file(s) during building

This what this could look like for the example: `say_hello.R` 

Support for `roxygen2` is built into `devtools`, so we can simply update the documentation from the command line like this:
```{r, eval = FALSE}
document()
load_all()
?say_hello
```

Note: text formatting, links, lists etc. work exactly as in `.Rd`, see \url{http://r-pkgs.had.co.nz/man.html#text-formatting}

### The `@export` tag

`roxygen2` solves the problem of maintaining separate `.Rd` files for documentation. 

`roxygen2` also solves the problem of keeping track of a separate NAMESPACE file: 

* only functions exported in NAMESPACE are directly avaiable when the package is loaded 
* by default, all functions defined in the package are exported via a call to `exportPattern`
* `document()` will add all functions with the `@export` tag to NAMESPACE
* functions without the tag will stay private (access through `:::`)

### Tips

Error messages like this are not uncommon when installing from `devtools`:

```
Error in fetch(key) : lazy-load database 'C:/Users/aleplo/Documents/R/R-3.6.1/library/say/help/say.rdb' is corrupt
```

Restart the R session to get rid of it. 

A `roxygen2`-style comment / documentation header is often useful when writing functions in scripts:

* Structured specification of what the function is supposed to do
* Easy upgrade to package down the road

## Writing good help

These are frequent recommendation for writing good documentation for R:

* Make the introduction short (1-5 lines)
* Avoid wordy argument descriptions, move extra material to Details section
* Describe the return value; make use of the list environment for structured return values (e.g. data frames)
* Use examples to demonstrate common use cases; make examples self-contained, using built-in datasets or simulated data; consider vignettes for complex use cases.

If you find it very hard to document your function(s), you may want to re-think your function- or package design. 
	
## Limitations of help

R-style documentation mixes reference and tutorial, but skews towards reference. It is often awkward for describing

* background and motivation, 
* multiple or complex use cases,
* use cases involving multiple functions 

Package-level documentation can help somewhat (see below), but generally these things are better handled via a vignette (or two).

# Vignettes

## Complement to `help`

* Generally: extra material in inst/doc in source package, copied over to install - could be anything (.pdf, .html, .docx (not recommended), paper, book chapter....)

* More specifically: a _package vignette_ is .pdf .HTML document in the built package obtained from plain text literate source files (.Rmd, .Rnw) in the source package (in inst/doc)

* Multiple vignettes possible, any content - but main idea: tutorial (as opposed to reference)

A good example (not format, but multiple vignettes with different use cases):
```{r, eval = FALSE}
browseVignettes(package = "survival")
```

* Unless it's literally a one-function package, have a vignette


## How to vignette

We can easily generate a vignette template for our silly example:

```{r, eval = FALSE}
use_vignette("SayingThings")
```

This creates a new sub-directory in the package folder called `vignettes` and puts a template `.Rmd` file there. 

You can edit this file as you see fit, though the current example may be a bit thin. This is what a an example could look like: `SayingThings.Rmd`

Vignettes are NOT built by default when using `install`, you need to insist:
```{r, eval = FALSE}
install(build_vignettes = TRUE)
browseVignettes(package = "say")
```

* Control output format by modifying the metadata of the vignette itself, i.e., use `rmarkdown::html_vignette` or `rmarkdown::pdf_document` as the output field (or anything else).

# Other 

## Adding package-level help

It can be helpful to have a central help page for the whole package. This can serve as a simple mini-vignette. E.g.:
```{r, eval = FALSE}
package?devtools
```

roxygen works for this, too:

1. Write documentation block for object `NULL`
2. Add tag `\@docType package` to the block
3. There is no step 3.

Common practice is to put this header into a separate `.R` file called `<package-name>.R.`, but any R file will do. 

This is what a package documentation for our package could look like: `say.R`

## Adding & documenting data sets

### Ready-made

If your package cannot use a standard data set for examples, include your own. 

Data sets live in directory `data` of the package folder. Technically, data can be stored as a text file, R code that will generate data, or an `.rda` file. The latter is most common and well supported by devtools:

```{r, eval = FALSE}
names = c("Ronald", "Herbert", "Bill", "George")
use_data( names )
```

generates file `data/names.rda` containing the single vector `names`. 

Data objects, too, need to be documented. This is again simplest using roxygen, by including the data name with a roxygen header in one of the package `.R` files. Here, it could be something like
```r
#' First names
#' 
#' A vector with four common English first names.
#' 
#' @format A character vector of length four. 
#' @source Synthetic data set
"names"
```
Documentation for data sets is built as before:
```{r, eval = FALSE}
document()
load_all()
?names
```

Comments: 

1. Command from Hadley: _Never \@export a data set._

2. You can use `\example`, too.


### Pre-processing files

Sometimes, you have an external data set that needs some clean up for inclusion in a package.

You could do this only once, at the command line, and  call `use_data`. However, for reasons of 

* documentation,
* replicability,
* maintenance

you really should do it in an R file & keep the R-file around. 

You could keep the .R file in `data`, in which case it will be run every time you load, built or install the package: that is a bit much.

His Hadleyness recommends an alternative workflow:

* Add a directory `data-raw` to the source package folder with code and raw data
* Exclude this directory from the build (`.Rbuildignore`)
* Run the code manually whenever the data needs updating

This is set up by calling `use_data_raw`. For our example:
```{r, eval = FALSE}
use_data_raw("corenames")
```

which sets up the directory, puts a script template with the given name there, updates `.Rbuildignore`, milks the cows and fixes the tractor. Your job is to copy the file name into the `data-raw` directory, write the preprocessing code, and run the script once to build the `.rda` file.

This is what a preprocessing script for data file `corenames.txt` could look like: `corenames.R`.

Of course, you still have to document the data...


# Exercises

1. Take the function `nestedCC` and add a roxygen header for documentation. Re-build the package, and check the help you have generated.

2. Add a simple vignette to the package with `nestedCC`.

3. Add the example file `cohort.txt` to the package:

    a. using `use_data`;
    b. using `use_data_raw`: clean up the data somewhat, e.g. define factors;
    c. write a short documentation of the data file.

4. Add documentation for the package itself.
    




